Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-01-06 Direction_JINST_v6.tex
version with referee comments and one pass by Henry. --Lindley

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
would have profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, for scintillator-based detectors, while the energy
resolution is good due to the abundance of light, the light is
isotropic and does not retain the directional information of the
primary particle.  In contrast, the direction of the particle can be
reconstructed from the Cherenkov cone in water-based detectors,
although the energy resolution rapidly degrades below $\sim$5~MeV. For
double-beta decay in particular, but also for neutrino interactions,
the directional information can be a strong suppressant of
backgrounds.

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double-beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application than the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.

@@

The absorption and emission spectra overlap at some level
in all molecules. Consequently, if there is only one type of molecule in
the scintillator cocktail the light output is reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluorophores are added to the
primary solvent to shift the wavelengths of the photons to higher values 
where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudocumene as the solvent with 1-5~g/l of
PPO as the fluorophore. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudocumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{d^2N}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. In a large detector,
the Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. These undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in figure \ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. As an alternative to single-atom-doping, recently nanocrystals 
formed by candidate isotopes have been explored as a dopant\cite{qdot,qdot2}.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}, for which digitized values come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure}

The studies in sections \ref{sim_section} to \ref{reconstruction_sec} are made with single 5~MeV electrons, lower energies are
discussed in section \ref{edep_section}. Future studies will work on increasing the complexity
of the event as is needed for $0\nu\beta\beta$ studies. Three
effects primarily contribute to the timing of the scintillator detector
system. First, the simulated travel time of an initial 5~MeV electron is typically between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) is a value which
can be achieved with the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD\cite{RSI_paper,PSEC4_paper,anode_paper}. We
note that uncertainties in the vertex reconstruction will produce a
similar effect to the smearing due to the TTS.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window. For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction, $\cos(\Psi) =
        x_{hit}/|\vec{r}_{hit}|$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. The $t\leq$ 34.0~ns cut is applied.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

Figure \ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

@@

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

@@

\section{Scintillator emission spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots; quantum dots can also provide a mechanism for introducing an 
isotope for studying double-beta decay.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34.0~ns timing cut.  Compared to
the default case shown in figure \ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are longer than in the KamLAND emission spectrum.

>>>>

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the $\sim$1-2~MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
would have profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe\cite{leptogenesis}.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\% \sigma_{e}$/\sqrt{E(MeV)}$\cite{kam2013,borexino}. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, this energy resolution is much poorer than other technologies being used for $0\nu\beta\beta$ searches: Ge detectors\cite{gerda2013}, Te bolometers\cite{Alessandria:2011rc}, tracking detectors\cite{SuperNEMO}, liquid Xe time projection chambers (TPCs)\cite{EXO2012} and high pressure gaseous Xe TPCs\cite{NEXTsipm}.

Scintillation light is isotropic. At these low energies, it does not contain sufficient to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

This is the first in a series of papers exploring directionality in large-scale liquid scintillator detectors. In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application from the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.  We then use these results as input into a traditional direction reconstruction developed for water Cherenkov detectors. Since the reconstruction of the direction of $\sim$1~MeV particles has not been achieved before, we start these studies with the most simple case of a single particle at the center of the detector. We also start with an easier test case of a 5~MeV electron like that from a $^{8}$B solar neutrino interaction. With the higher photon statistics at this energy, we verify the technique and the different detector parameters that affect it. We then lower the energy to that relevant to a single electron in a $0\nu\beta\beta$.

@@

The absorption and emission spectra overlap at some level
in all molecules. Consequently, if there is only one type of molecule in
the scintillator cocktail the light output is reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluorophores are added to the
primary solvent to shift the wavelengths of the photons to higher values 
where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudocumene(1,2,4-trimethylbenzene) as the solvent with 1-5~g/l of
PPO (2,5-diphenyloxazole) as the fluorophore. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudocumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{d^2N}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $N$ is the number of Cherenkov photons, $\lambda$ is the wavelength of the photon, x is the distance travelled by the particle, $Z$ is the charge of the particle, $\alpha$ is the fine structure constant, $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the particle.

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. For the example scintillator above, n(400nm)=1.459\cite{OlegThesis}. The absorption cutoff is between 370~nm and 360~nm where the attenuation length rapidly drops from 6.5~m to 0.65~m. For a 5~MeV electron, this yields 744 Cherenkov photons per event; dropping down to a 1~MeV electron yields 89 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm, the point above which a small number of photons is detected due to a standard photocathode's quantum efficiency. For reference at this cutoff wavelength, the index of refraction n(370nm)=1.466. This can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV.

All photons including these undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
This means photons at the scintillation cutoff at 370~nm will travel at 0.191~m/ns while those at 600~nm travel at 0.203 m/ns. Since on average the undisturbed Cherenkov photons have longer wavelengths, they will arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}.
For light Majorana neutrino exchange, a probable case is the equal division of energy between two
electrons emitted at a large angle relative to each other\cite{phasespace, SuperNEMO}. An example case is shown in figure \ref{detector_view}
for a $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred. Due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains, isotopes with Q-values above the 2.6~MeV gamma from $^{208}$Tl are preferred; however slightly lower Q-values can be tolerated if there is sufficient background rejection. There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with sufficiently large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently nanocrystals formed by candidate isotopes have been explored as an alternative to doping by single atoms,\cite{qdot,qdot2}.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The most critical value is the attenuation length at 400~nm which is 25~m. The attenuation length drops precipitously at 370~nm from 6.5~m to 0.65~m at the cutoff wavelength at 360~nm.
Variations from the baseline KamLAND case are discussed below.

Re-emission of absorbed photons in the scintillator
bulk volume and optical scattering, specifically Rayleigh scattering, have not yet been included by default. A test simulation shows that the effect of optical scattering is negligible. The reason is that the cutoff is very steep below 360~nm and almost no photons reach the sphere, so optical scattering makes no difference for short wavelengths. Above about 395~nm, the attenuation length is greater than 20~m so both scattering and absorption are not very likely and scattering is negligible. The intermediate region is rather small. A similar argument holds for re-emission. Scattering length measurements and discussions can be found in Ref.~\cite{Wurm:2010ad}.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Similar to optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~EPS
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

%The studies in sections \ref{sim_section} to \ref{reconstruction_sec} are made with single 5~MeV electrons, lower energies are
%discussed in section \ref{edep_section}. Future studies will work on increasing the complexity
%of the event as is needed for $0\nu\beta\beta$ studies.

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however before the electron drops below Cherenkov threshold the scattering angle is small and therefore the Cherenkov light encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

The scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. The first step in the scintillation process is the transfer of energy from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator.

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences due to this chromatic dispersion, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The higher energy provides larger photon statistics with which to evaluate the method, debug the simulation and is of interest to neutrino-electron scattering experiments. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) can be achieved with large area picosecond photodetectors
(LAPPDs)\cite{Adams:2013nva,RSI_paper,PSEC4_paper,anode_paper} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD\cite{RSI_paper,PSEC4_paper,anode_paper}.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window, see figure \ref{time_plots_comparison} (a). For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs2/combination_cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction, $\cos(\Psi) =
        x_{hit}/|\vec{r}_{hit}|$. Three energies are shown: 5~MeV (Black - Top), 2.1~MeV (Red - Middle), 1.4~MeV (Blue - Bottom). Each sample consists of 1000 events produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. The $t\leq$ 34.0~ns cut is applied.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

Figure \ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial electrons even at 1.4~MeV.

@@

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons that pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector that is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

@@

\section{Scintillator emission spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectrum by tuning the size or composition of the quantum
dots.

Quantum dots can also provide a mechanism for introducing an 
isotope for studying double-beta decay. The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34.0~ns timing cut.  Compared to
the default case shown in figure \ref{time_plots_comparison}(a), the
separation is worse (as expected) because the scintillation light
wavelengths are longer than in the KamLAND emission spectrum.
